---
title: 'acs14lite: A lightweight R interface to the 2010-2014 ACS API'
author: "Kyle Walker"
date: 2015-12-12
slug: acs14lite
tags: ["R"]
---

I use data from the US Census Bureau's American Community Survey all of the time.  I also use R all of the time.  Naturally, this means that I often use ACS data in R - which is pertinent given last week's release of the new 2010-2014 ACS estimates.  I wanted easy access to the data to facilitate my on-going research on demographic trends in US metros, and work at the TCU Center for Urban Studies; as such, I wrote a small R package to provide quick access to the data, __acs14lite__ (https://github.com/walkerke/acs14lite).  This is not intended to be comparable to, or a replacement for, the existing ACS package in R; it is more for my personal convenience, but I thought it might be useful to others as well.  This is mostly going to be a side project for me, so I don't have plans for a CRAN submission at this time.  

Install from GitHub with the following command in R: 

```{r, eval = FALSE}

devtools::install_github('walkerke/acs14lite')

```

Accessing the US Census Bureau's API requires an API key, which you can get from here: http://api.census.gov/data/key_signup.html.  You can then set it globally in your __acs14lite__ session: 

```{r, eval = FALSE}
library(acs14lite)

set_api_key('your API key here')

```

```{r, echo = FALSE}
library(acs14lite)

set_api_key("5ed58a5745802102fb83d4eec5d1f7326f65ffab")
```

There is one main function in the package: `acs14`.  From here, you can request data for the following geographies: the entire US, regions, divisions, states, counties, Census tracts, and Census block groups.  These are the geographies that I generally use, and I don't have plans at the moment to add more; I would welcome pull requests, however.  

The `acs14` function has the following parameters: 

* `api_key`: If you've set your API key already with `set_api_key`, you don't need to provide this.  
* `geography`: One of 'us' (the default), 'region', 'division', 'state', 'county', 'tract', or 'block group'.  
* `variable`: A character string representing the Census variable name you want, or a vector of multiple variable names.  Defaults to 'B01001_001E', which is total population.  You can use the ACS package to look for variable names with its `acs.lookup` function; remember to add `E` for estimate and `M` for margin of error to the end of your variable name.  
* `state`: The name of the state for which you want data; applicable to counties, tracts, and block groups.  
* `county`: The name of the county for which you want data: applicable to tracts and block groups.  

The function returns an R data frame with the data you want for your requested geography.  

Additionally, I've written a few functions to help users work with margins of error in the ACS.  Margins of error for the raw data are provided from the API; however, we often calculate new variables based on the ACS estimates, which in turn will have their own respective margins of error.  I've used the guidelines in Appendix 3 here: https://www.census.gov/content/dam/Census/library/publications/2008/acs/ACSGeneralHandbook.pdf to write the following functions: 

* `moe_sum`: calculates a margin of error for a derived sum of ACS estimates
* `moe_prop`: calculates a margin of error for a proportion
* `moe_ratio`: calculates a margin of error for a ratio
* `moe_product`: calculates a margin of error for a product

Below, I provide a couple examples of how you can use the package.  

### Interactive dot plot of income by county in Wyoming with Plotly

```{r, warning = FALSE, message = FALSE}
library(ggplot2)
library(plotly)
library(dplyr)

wy_income <- acs14(geography = 'county', variable = c('B19013_001E', 'B19013_001M'), state = 'WY')

wy2 <- wy_income %>%
  mutate(name = gsub(" County, Wyoming", "", wy_income$NAME),
         low = B19013_001E - B19013_001M,
         high = B19013_001E + B19013_001M) %>%
  select(name, low, high, estimate = B19013_001E) %>%
  arrange(desc(estimate))

g <- ggplot(wy2, aes(x = estimate, y = reorder(name, estimate))) +
  geom_point() +
  geom_errorbarh(aes(xmin = low, xmax = high)) +
  xlab("Median household income, 2010-2014 ACS estimate") +
  ylab("")


ggplotly(g) %>% layout(margin = list(l = 120))
```

### Interactive map of poverty in Los Angeles County by Census tract with CartoDB and the __tigris__ package

```{r, eval = FALSE}
library(tigris)
library(CartoDB) # devtools::install_github("becarioprecario/cartodb-r/CartoDB", dep = TRUE)
library(rgdal)

la_poverty <- acs14(geography = 'tract', state = 'CA', county = 'Los Angeles',
                    variable = c('B17001_001E', 'B17001_001M', 'B17001_002E', 'B17001_002M'))

la2 <- la_poverty %>%
  mutate(geoid = paste0(state, county, tract),
         pctpov = round(100 * (B17001_002E / B17001_001E), 1),
         moepov = round(100 * (moe_prop(B17001_002E, B17001_001E, B17001_002M, B17001_001M)), 1)) %>%
  select(geoid, pctpov, moepov)

cdb_name <- 'your CartoDB username here'
cdb_key <- 'your CartoDB API key here'

cartodb(cdb_name, cdb_key)

la_tracts <- tracts('CA', 'Los Angeles', cb = TRUE)

la_tracts2 <- geo_join(la_tracts, la2, "GEOID", "geoid")

r2cartodb(la_tracts2, 'la_poverty')

# Now, head to your CartoDB account to style your map!
```

<iframe width="100%" height="520" frameborder="0" src="https://kwalkertcu.cartodb.com/viz/2d99c916-a293-11e5-835d-0e5db1731f59/embed_map" allowfullscreen webkitallowfullscreen mozallowfullscreen oallowfullscreen msallowfullscreen></iframe>

