---
title: "Dropdown menus in R Markdown with bsselectR"
author: "Kyle Walker"
date: 2016-12-15
tags: ["R", "presentations", "R Markdown"]
slug: rmd-dropdowns
---

I strongly believe that interactive reports, presentations, and scholarly articles are going to become much more prominent in the years ahead.  Whereas a PDF article or presentation can often only show a limited aspect of a research project, interactive documents can allow a reader or presenter to explore project content in a much broader sense.  

For dynamic research documents, an excellent option is to combine Shiny with R Markdown to generate a report that can execute R code from a Shiny server.  However, not all interactive documents need the full power of Shiny.  For example, I'm currently writing a report in which I'd like readers to be able to use drop-down menus to select charts to view.  As I produced the charts in Python and saved them all to a folder, I don't need Shiny to dynamically create the charts in the report.  

To create the drop-down menu, I came across the [bootstrap-select](https://silviomoreto.github.io/bootstrap-select/) jQuery plugin.  bootstrap-select adds Bootstrap's dropdown.js and Bootstrap styling to re-style HTML `<select>` menus, which makes it a nice fit for R Markdown documents that default to this styling.  While writing out the raw HTML and JavaScript to accomplish this worked in R Markdown, I found this tedious, so I decided to author __bsselectR__, an htmlwidget to do this automatically with R code.  Many, many thanks are due to the developers of bootstrap-select for their hard work and excellent documentation.  

bsselectR is not on CRAN, but you can install the development version from GitHub with `devtools::install_github("walkerke/bsselectR")`.  View the source code at https://github.com/walkerke/bsselectR.  

At the moment, there is one main function in bsselectR, `bsselect`, which will generate a bootstrap-select dropdown menu that allows you to interactively display text, images, or HTML documents as iframes.  Pass a named vector of text or paths to images/HTML documents and specify the type as `"text"`, `"img"`, or `"iframe"`, and `bsselect` will return a drop-down menu.  On a basic level, `bsselect` can work outside of R Markdown, using the default Bootstrap styling.  Let's try a simple example: 

```{r eval = FALSE}
library(bsselectR)

quotes <- c("Look deep into nature, and then you will understand everything better.", 
            "A fool thinks himself to be wise, but a wise man knows himself to be a fool.", 
            "My mission in life is not merely to survive, but to thrive; and to do so with some passion, some compassion, some humor, and some style.")

names(quotes) <- c("Einstein", "Shakespeare", "Angelou")

bsselect(quotes, type = "text")
```

<iframe src="https://walkerke.github.io/extras/quotes.html" height="150" width = "750" frameborder = "0" scrolling = "no"></iframe>

Use the menu to flip through the three quotes.  

A more practical example, however, would be the use of `bsselect` in an R Markdown document where the user could flip through a number of plots that are stored in a subdirectory.  Let's take a recent, applied example: scatterplots of the relationship between the percentage vote for Donald Trump and the percentage of county residents with at least a bachelor's degree by state.  

We'll get the election data from [Mike Kearney's open presidential election dataset](https://github.com/mkearney/presidential_election_county_results_2016) (updated December 9, so there may be some undercounts) and education data from the ACS using my personal [acs14lite package](https://github.com/walkerke/acs14lite).  The following code will generate identical plots for each state, and save them in a subdirectory of my working directory named `plots`.  

```{r, eval = FALSE}
library(tidyverse)
library(scales)
library(acs14lite)

votes16 <- read_csv("https://raw.githubusercontent.com/mkearney/presidential_election_county_results_2016/master/pres.elect16.results.dec9.csv")

vars_to_get <- c("B15002_001E", "B15002_015E", "B15002_016E",  "B15002_017E",  "B15002_018E",
                 "B15002_032E",  "B15002_033E",  "B15002_034E",  "B15002_035E", "B01001_001E")

educ <- acs14(api_key = "Your key goes here", 
              geography = "county", variable = vars_to_get)

educ <- educ %>%
  mutate(pctba = (B15002_015E + B15002_016E + B15002_017E + B15002_018E + B15002_032E + 
                    B15002_033E + B15002_034E + B15002_035E)  / B15002_001E, 
         fips = paste0(state, county))

state_match <- data.frame(state_name = state.name, st = state.abb)

df <- votes16 %>%
  mutate(fips = stringr::str_pad(fips, 5, "left", "0")) %>%
  filter(cand == "Donald Trump") %>%
  inner_join(educ, by = "fips") %>%
  left_join(state_match, by = "st")

states <- as.character(sort(unique(df$state_name)))
states <- states[!is.na(states)]
states <- states[!states %in% c("District of Columbia", "Alaska")]

plots <- lapply(states, function(x) {
  
  state = filter(df, state_name == x)
  
  p <- ggplot(state, aes(x = pctba, 
              y = pct)) +
    geom_point(aes(size = B01001_001E), alpha = 0.5, color = "blue") + 
    geom_smooth(method = "lm") + 
    scale_y_continuous(labels = percent_format()) + 
    scale_x_continuous(labels = percent_format()) + 
    scale_size_continuous(labels = comma_format()) + 
    theme_minimal() + 
    labs(title = x, 
         x = "Percent over age 25 with at least a bachelor's degree", 
         y = "Percent voting Trump", 
         size = "Total population", 
         caption = "Data sources: Mike Kearney, 2010-2014 ACS")
  
  p
  
})

paths <- paste0(states, ".png")

pwalk(list(paths, plots), ggsave, path = "plots", dpi = 300)
```

Now, the following code in an R Markdown document knit from my working directory will get me interactivity without Shiny!  I've also set the selected option to "Oregon", and the parameters `live_search` and `show_tick` to `TRUE`, which allows for interactive searching of the drop-down menu and shows a check mark next to the selected option.  The bootstrap-select dropdown adopts the theme of the R Markdown document, which in this case is `spacelab`.  

```{r, eval = FALSE}
library(stringr)
library(bsselectR)

state_plots <- paste0(list.files("plots", full.names = TRUE))
names(state_plots) <- str_replace_all(state_plots, 
                                      c("\\.png" = "", 
                                        "plots/" = ""))

bsselect(state_plots, type = "img", selected = "Oregon", 
         live_search = TRUE, show_tick = TRUE)
```

<iframe src="https://walkerke.github.io/extras/trump_by_education.html" width="100%" height = "825" scrolling="no" frameborder="0"></iframe>

Publish your directory to your hosting service of choice, and you'll have an interactive report to share like the one above.  

`bsselect` also works in other R Markdown documents, such as presentations!  The example below shows a bsselectR dropdown in a [xaringan](https://github.com/yihui/xaringan) [remark.js](https://remarkjs.com/#1) presentation.  To avoid CSS/JS conflicts, I've wrapped the call to `bsselect` in bsselectR's `as_iframe` function, which will save out the widget and embed it as an iframe.  

<iframe src="https://walkerke.github.io/extras/xaringan.html" width="100%" height="600" frameborder = "0"></iframe>

I'd welcome feedback and pull requests; I'll be adding features to the package bit by bit in the time ahead.  


