---
title: "Fetching roads data in R with tigris"
author: "Kyle Walker"
date: 2015-12-03
slug: roads-tigris
tags: ["R"]
---

There are three functions available in the __tigris__ package (https://github.com/walkerke/tigris) to fetch road data.  `primary_roads()` loads all interstates for the entire US; `primary_secondary_roads()` gets you interstates and US/state/county highways, by state; and `roads()` gets you all road segments for a given county within a state.  In this example, we'll use the `primary_secondary_roads()` function to get our data for Route 1 in California.  

```{r, message = FALSE, warning = FALSE}

library(tigris)
library(leaflet)
library(rgdal)
library(geojsonio)
library(widgetframe)

ca <- primary_secondary_roads(state = 'California')

rt1 <- ca[ca$FULLNAME == 'State Rte 1', ]

```

We can then plot with the `leaflet` package: 

```{r, warning = FALSE}

map <- leaflet(rt1) %>% addProviderTiles('CartoDB.Positron') %>% addPolylines()

frameWidget(map)

```

Using the `geojsonio` package, we can then write to GeoJSON for use in other applications.  Before doing this, I'd advise transforming the coordinate system to WGS84 from NAD83, which is used by all of the Census shapefiles; the two are functionally equivalent, but WGS84 is more universally recognized (e.g. on GitHub).  

```{r, eval = FALSE}

rt1 <- spTransform(rt1, CRS("+proj=longlat +datum=WGS84"))

geojson_write(rt1, file = 'route1.geojson')

```

You can view the resultant GeoJSON as a GitHub Gist here: https://gist.github.com/walkerke/c3501f481a780834f8e8

I should note that the `primary_secondary_roads()` function returns an R object of class `SpatialLinesDataFrame` with 177 different line segments that collectively make up Route 1.  If need be, these segments can be combined with the `gLinesMerge` function in the `rgeos` package.  
