---
title: "Compare US metropolitan area characteristics in R with tidycensus and tigris"
author: "Kyle Walker"
date: 2017-06-25T08:00:00
tags: ["r", "census", "tidycensus", "tigris"]
slug: comparing-metros
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, out.width = "100%")
library(tidycensus)
census_api_key("5ed58a5745802102fb83d4eec5d1f7326f65ffab")
```

* Interested in more tips on working with Census data? [Click here to join my email list!](http://eepurl.com/cPGKZD)
* Want to implement this in your organization? Contact me at kwalkerdata@gmail.com to discuss a training or consulting partnership.  

As I've discussed [in a previous post](https://walkerke.github.io/2017/05/tigris-metros/), practioners commonly analyze demographic or economic topics at the scale of the metropolitan area.  Since I wrote that post, I've released the [tidycensus](https://walkerke.github.io/tidycensus) package, giving R users access to linked Census geometry and attributes in a single function call.  This makes metropolitan area analysis even faster, with help from the __tigris__ and __sf__ packages.    

First, we load up some packages and set some options.  


```{r}
library(tidycensus)
library(tidyverse)
library(tigris)
library(sf)
options(tigris_class = "sf")
options(tigris_use_cache = TRUE)
# census_api_key("YOUR KEY HERE")
```


Let's say you are an analyst who wants to compare the distribution of median gross rent by Census tract for three West Coast metropolitan areas: Seattle, Portland, and San Francisco-Oakland.  We'll get data from the 2011-2015 American Community Survey's Data Profile, using variable `DP04_0134`.  As this requires data for multiple states, we'll use a new feature in tidycensus that allows us to supply a vector of states to the `state` parameter in `get_acs()` for Census tracts, improving on the approach I outlined [in my previous post](http://walkerke.github.io/2017/05/tidycensus-every-tract/).  

```{r}
rent <- get_acs(geography = "tract", variables = "DP04_0134", 
                state = c("WA", "OR", "CA"), geometry = TRUE)

head(rent)

```


We now have median gross rent information for all Census tracts in Washington, Oregon, and California.  Subsetting this to our desired metropolitan areas only takes a couple steps.  First, we use the tigris package to obtain metropolitan area boundaries with the `core_based_statistical_areas()` function, taking care to set `cb = TRUE` as this is the default geometry used by tidycensus, and subset by ID for our desired metros.  Next, we use an inner spatial join with sf's `st_join()` function to subset for the specific tracts we need.  

```{r}
metros <- core_based_statistical_areas(cb = TRUE) %>%
  filter(GEOID %in% c("38900", "41860", "42660")) %>%
  select(metro_name = NAME)

wc_rent <- st_join(rent, metros, join = st_within, 
                   left = FALSE) 

head(wc_rent)

```

Tracts are identified by metropolitan area, with a new column, `metro_name`, that includes the metropolitan area name.  

Exploratory analysis by metropolitan area is now straightforward.  We can look at faceted histograms of median gross rent by Census tract by metro area with __ggplot2__: 

```{r}
ggplot(wc_rent, aes(x = estimate)) + 
  geom_histogram() + 
  facet_wrap(~metro_name)

```

As we have feature geometry as well, we can make faceted maps with `geom_sf()`, found in the development version of ggplot2.  Be sure to set `scales = "free"` and `theme(aspect.ratio = 1)` if you want this to work correctly.   

```{r, fig.width = 10, fig.height = 4}
library(viridis)

ggplot(wc_rent, aes(fill = estimate, color = estimate)) + 
  geom_sf() + 
  coord_sf(crs = 26910) + 
  facet_wrap(~metro_name, scales = "free", nrow = 1) + 
  theme_minimal() + 
  theme(aspect.ratio = 1) + 
  scale_fill_viridis() + 
  scale_color_viridis()
  
```


